# Reproducible Research: Peer Assessment 1

### Course information  
Coursera: Reproducible Research  
Professors: Roger D. Peng, PhD; Jeff Leek, PhD; Brian Caffo, PhD  
[Course website][courselink]  

### Data  
The data for this assignment can be downloaded from the course web site:  
* Dataset: [Activity monitoring data][datalink] (52K)

The variables included in this dataset are:  
* **steps**: Number of steps taking in a 5-minute interval (missing values are coded as `NA`)  
* **date**: The date on which the measurement was taken in YYYY-MM-DD format  
* **interval**: Identifier for the 5-minute interval in which measurement was taken  

The dataset is stored in a comma-separated-value (CSV) file and there are a total of 17,568 observations in this dataset.  

## Loading and preprocessing the data
Read in data file:
```{r}
data <- read.csv("activity.csv")
```

Convert date values to Date type:
```{r}
data$date <- as.Date(data$date, "%Y-%m-%d")
```

## What is mean total number of steps taken per day?
Calculate the number of steps taken each day:
```{r}
sum_per_day <- aggregate(steps ~ date, data=data, sum)
```

Make a histogram of the total number of steps taken each day:
```{r}
hist(sum_per_day$steps, xlab="Number of Steps Per Day", 
     main="Histogram of Number of Steps Per Day")
```

Calculate and report the **mean** and **median** total number of steps taken per day:
```{r}
mean_per_day <- round(mean(sum_per_day$steps))
median_per_day <- round(median(sum_per_day$steps))
```

The **mean** total number of steps per day is `r mean_per_day`.  
The **median** total number of steps per day is `r median_per_day`.

## What is the average daily activity pattern?
Calculate the mean number of steps taken during each interval:  
```{r}
mean_per_interval <- aggregate(steps ~ interval, data=data, mean)
```

Make a time series plot (i.e. `type = "l"`) of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all days (y-axis).  
```{r}
plot(mean_per_interval$interval, mean_per_interval$steps, type="l",
     xlab="Interval", ylab="Mean Number of Steps", 
     main="Average Daily Activity Pattern")
```

Which 5-minute interval, on average across all the days in the dataset, contains the maximum number of steps?  
```{r}
max_interval <- mean_per_interval$steps==max(mean_per_interval$steps)
mean_per_interval[max_interval,]$interval
```


## Imputing missing values
Calculate and report the total number of missing values in the dataset (i.e. the total number of rows with `NA`s):
```{r}
numNA <- sum(is.na(data$steps))
numNA
```

Devise a strategy for filling in all of the missing values in the dataset. The strategy does not need to be sophisticated. For example, you could use the mean/median for that day, or the mean for that 5-minute interval, etc.

Fillin `NA` values with mean for that 5-minute interval:  
```{r}
## create a list of rows with data to impute
to_impute <- grep(TRUE, is.na(data$steps))

##copy "data" dataframe to "fillin_data" dataframe
fillin_data <- data

##for each index in impute list, replace the NA value with the mean
## number of steps for that interval from the mean_per_interval dataframe
for (i in to_impute){
    fillin_data[i,]$steps <- mean_per_interval[(mean_per_interval$interval == 
                                               fillin_data[i,]$interval),]$steps
}
```

Calculate the number of steps taken each day:
```{r}
sum_per_day_fillin <- aggregate(steps ~ date, data=fillin_data, sum)
```

Make a histogram of the total number of steps taken each day:
```{r}
hist(sum_per_day_fillin$steps, xlab="Number of Steps Per Day", 
     main="Histogram of Number of Steps Per Day (Fill-in Data)")
```

Calculate and report the **mean** and **median** total number of steps taken per day:
```{r}
mean_per_day_fillin <- round(mean(sum_per_day_fillin$steps))
median_per_day_fillin <- round(median(sum_per_day_fillin$steps))
```

The **mean** total number of steps per day is `r mean_per_day_fillin`.  
The **median** total number of steps per day is `r median_per_day_fillin`.

Do these values differ from the estimates from the first part of the assignment?  
**The median value for the fill-in data has increased by 1 step per day.**  
Please see summary table below:
```{r kable, results="asis", echo=FALSE}
summary1 <- data.frame(dataset=c("data", "fillin_data"), 
            mean=c(mean_per_day, mean_per_day_fillin), 
            median=c(median_per_day, median_per_day_fillin))
kable(summary1, digits = 1)
```

What is the impact of imputing missing data on the estimates of the total daily number of steps? 
```{r}
ttest <- t.test(sum_per_day$steps, sum_per_day_fillin$steps)
ttest$p.value
```
There is no statistically significant impact on the estimates of the total daily number of steps.


## Are there differences in activity patterns between weekdays and weekends?

Create a new factor variable in the dataset with two levels – “weekday” and “weekend” indicating whether a given date is a weekday or weekend day.  
```{r}
fillin_data$weekday <- weekdays(fillin_data$date)
fillin_data$group <- "weekday"
filter <- grep("^S", fillin_data$weekday)
fillin_data[filter,]$group <- "weekend"
```
Make a panel plot containing a time series plot (i.e. `type = "l"`) of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all weekday days or weekend days (y-axis).

```{r}
facet_plot_data <- aggregate(steps ~ group + interval, data=fillin_data, mean)

library(ggplot2)
qplot(interval, steps, data=facet_plot_data, geom="line", facets=group ~ .)
```

[datalink]: https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip
[courselink]: https://class.coursera.org/repdata-004/